---
title: "White Wine quality Analysis"
output: html_document

---

This is an attempt to explore the various properties that affect the quality 
score of White Wine. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, messages=FALSE)
library(ggplot2)
library(GGally)
library(dplyr)
library(reshape2)
library(gridExtra)
```

Set Working Directory and Load the Input csv. 

```{r setwd}
setwd("C:/udacity/EDA_USING_R/project")
whiteWineData<-read.csv("wineQualityWhites.csv")
#qplot(quality, data = whiteWineData, fill = color, binwidth = 1) +
#    scale_x_continuous(breaks = seq(3,10,1), lim = c(3,10)) +
#    scale_y_sqrt()
```
### Summary of the Data
```{r Summary}
summary(whiteWineData)
#whiteWineData$color<-"white"
```

From the summary , some of the median values are lesser than the Mean values. 
Possible outliers or data skew. Also the Max value of free sulfur.dioxide and 
residual.sugar are way beyond the other values. Otherwise most of them seem like
normal distributions.But lets plot to verify .

```{r Structure_and_names}
str(whiteWineData)
names(whiteWineData)
```

### Points of interest that will be explored
1. Which input variables have high correlation with the output variable quality
2. Do input variables have any relation or are they independent (for e.g impact 
of pH on alcohol, residual sugars on density etc.,)

I want to start by checking the correlation between Quality and other inputs to 
see if there are any strong candidates that we can start our analysis with.

```{r inputs_quality_correlation}
cor(x=whiteWineData[,2:12], y=whiteWineData$quality,method="pearson")
```

From the values ,none of them have a very strong correlation but we can see that
alcohol has a good positive correlation.pH and sulphates have a positive but 
week correlation. Density seems to have a relatively high negative correlation 
followed by chlorides , volatile.acidity and total.sulfur.dioxide

Below are a few utility functions that will be used later during the analysis.

##Utility functions
 Below function f_qualitygroup is used to return the dataset after grouping by 
 the grouping  variable.
 Although we need only to group by quality as part of this analysis , we can use 
 any variable passed as grouping_var and input_var
 
```{r}

library(dplyr)
library(lazyeval)

f_qualitygroup <- function(data,grouping_var,input_var){

 data.quality_group<-data %>%
  group_by_(grouping_var) %>%
  summarise_(group_mean=interp(~mean(input_var),input_var=as.name(input_var)),
            group_median=interp(~median(input_var),input_var=as.name(input_var)),
            n=~n()
            )
 return (data.quality_group)

}

```

Function to draw scatterplot with a few options needed for the analysis. It includes the Linear regression plot as well as the mean as the hline 

```{r scatterplot_function}

f_drawscatter <- function(dataset, x, y,  opts=NULL) {
  ggplot(dataset, aes_string(x = x, y = y)) +
   geom_jitter(alpha = 1/5) +
   geom_smooth(method = 'lm')+
    geom_hline(yintercept = mean(dataset[,y]),linetype='longdash',alpha=0.5,
               color="blue")
}

```

Function to draw scatterplot with similar options as the scatterplot.

```{r boxplot_function}

f_drawbox <- function(dataset, x, y,  opts=NULL) {
  ggplot(dataset, aes_string(x = x, y = y)) +
   geom_boxplot(alpha = 1/5) +
   geom_smooth(method = 'lm')+
    geom_hline(yintercept = mean(dataset[,y]),linetype='longdash',alpha=0.5,
               color="blue")
}

```

Let us analyze each variable  starting with distribution of Quality 

##Univariate  and BiVariate analysis

### 1. Quality

```{r Quality_sumary}
summary(whiteWineData$quality)
table(whiteWineData$quality)
```

Quality has a fixed set of values in the dataset.
From the table values , we can see that around 20 of them have the lowest 
quality of 3 and 5 have the highest quality rating of 9. Almost 60% of the 
samples have Quality rating 5 and 6 .

Let us plot a histogram.

```{r}
quality_plot=ggplot(data=whiteWineData,aes(x=quality))
quality_plot+geom_histogram(binwidth=1,color=I('black'),fill=I('#099009'),origin=-0.5)+
  scale_x_continuous(breaks = seq(2,10,1), lim = c(2,10))
```

From the plot , we can see that lot of wines have quality between 5 to 8 and 
there is none below 3 and above 9.

### 2. Alcohol 

```{r alcohol_summary}
summary(whiteWineData$alcohol)
```

Mean is slightly higher than Median and Max value is 14.2.There must be some 
skew in the data.

```{r alcohol_plots}
alcohol_plot=ggplot(data=whiteWineData,aes(x=alcohol))
alc_hist<-alcohol_plot+geom_histogram()
alc_dens<-alcohol_plot+geom_density()
grid.arrange(alc_hist,alc_dens,nrow=2)
```

The dataset looks a little positively skewed. Alcohol levels are between 8 and 14.


Let us study the affect of alcohol on quality .

Let us put the grouping utility function to use on quality and alcohol

### 3. Quality vs Alcohol

'##' quality_alcohol_group contains the mean and median values of the alcohol content in each '##'quality group
```{r quality_alcohol_grouping}
whiteWineData.quality_alcohol_group<-f_qualitygroup(whiteWineData,"quality"
                                                    ,"alcohol")
summary(whiteWineData$alcohol)
head(whiteWineData.quality_alcohol_group)
```

From the above grouping , we can see that the wines that have quality score of 
6,7,8 have mean alcohol levels greater than the overall mean alcohol level 
of 10.5 . Interestingly the ones that have quality score of 3,4  have mean 
alcohol values close to the overall Mean and greater than the ones with quality 
score of 5. There may be other factors affecting the score in case of these 
which shadowed alcohol quantity.

Let us draw plots that depicts these above numbers.

```{r quality_vs_alcohol}
qual_alc_scatterplot <-f_drawscatter(whiteWineData,"quality","alcohol")
qual_alc_boxplot <-f_drawbox(whiteWineData,"factor(quality)","alcohol")
grid.arrange(qual_alc_scatterplot,qual_alc_boxplot)

```

From the plots, we can see that there is a linear positive pattern between 
quality and alcohol level but it is also evident that there are other factors 
alongside alcohol that might have affected the quality.

### 4. Density

Let us focus on density vs quality .

```{r density_sumary}
summary(whiteWineData$density)
```

Mean is slightly higher than median and probably due to the data skew

```{r density_plots}
density_plot=ggplot(data=whiteWineData,aes(x=density))+
  scale_x_continuous(limits=c(0.98,1.01))
dens_hist<-density_plot+geom_histogram()
winedens_dens<-density_plot+geom_density()
grid.arrange(dens_hist,winedens_dens,nrow=2)
```

There are two odd peaks and data seems to be a little right skewed.

###5 .Density vs Quality

'##' similar to alcohol, let us create a variable to hold mean and median values of density for '##'each quality group.
```{r quality_vs_density}
whiteWineData.quality_density_group<-f_qualitygroup(whiteWineData,"quality",
                                                    "density")
summary(whiteWineData$density)
head(whiteWineData.quality_density_group)

```


```{r quality_vs_density_box}
qual_den_scatterplot <-f_drawscatter(whiteWineData,"quality","density")+
  scale_y_continuous(limits=c(0.985,1.005))
qual_den_boxplot <-f_drawbox(whiteWineData,"factor(quality)","density")+
  scale_y_continuous(limits=c(0.985,1.005))
grid.arrange(qual_den_scatterplot,qual_den_boxplot)

```

From the plots , we can see that the ones with high quality rating have density lower than the mean overall density across samples and the quality increased with a decrease in density.

Let us check the effect of chlorides on quality


###6 . Chlorides 

```{r}

chlorides_plot=ggplot(data=whiteWineData,aes(x=chlorides))+scale_x_continuous(limits=c(0.001,0.1))
chlor_hist<-chlorides_plot+geom_histogram()
chlor_dens<-chlorides_plot+geom_density()
grid.arrange(chlor_hist,chlor_dens,nrow=2)

```

Seems like a normal distribution except for the twin peaks.

###7 . Quality vs Chlorides

```{r quality_vs_chlorides}
whiteWineData.quality_chlorides_group<-f_qualitygroup(whiteWineData,"quality",
                                                      "chlorides")
summary(whiteWineData$chlorides)
head(whiteWineData.quality_chlorides_group)

```

The ones with high quality score seem to have less amount of chlorides compared 
to the low quality score ones

```{r quality_vs_chlorides_plots}
qual_chl_scatterplot <-f_drawscatter(whiteWineData,"quality","chlorides")+
  scale_y_continuous(limits=c(0,0.2))
qual_chl_boxplot <-f_drawbox(whiteWineData,"factor(quality)","chlorides")+
  scale_y_continuous(limits=c(0,0.2))
grid.arrange(qual_chl_scatterplot,qual_chl_boxplot)
```


From the plots , we can see that the ones with high quality rating have 
chlorides lower than the mean overall chlorides across samples.

Let us also check the relation between total.sulfur.dioxide and quality.

### 8. Total SO2 

```{r}
totalso2_plot=ggplot(data=whiteWineData,aes(x=total.sulfur.dioxide))+
  scale_x_continuous(limits=c(0,300))
totalso2_hist<-totalso2_plot+geom_histogram()
totalso2_dens<-totalso2_plot+geom_density()
grid.arrange(totalso2_hist,totalso2_dens,nrow=2)

```

Seems like a normal but a slight right skew in the disribution.

### 9. Total SO2  vs Quality

```{r quality_vs_totalso2}
whiteWineData.quality_totalso2_group<-f_qualitygroup(whiteWineData,
                                                     "quality","total.sulfur.dioxide")
summary(whiteWineData$total.sulfur.dioxide)
head(whiteWineData.quality_totalso2_group)

```

From the numbers , except for the wines with quality score 4 , there is a negative linear relation between quality and total.sulfur.dixoide. Let us plot to confirm.

```{r quality_vs_totalso2 plots}
qual_so2_scatterplot <-f_drawscatter(whiteWineData,"quality",
                                    "total.sulfur.dioxide")+
  scale_y_continuous(limits=c(0,300))
qual_so2_boxplot <-f_drawbox(whiteWineData,"factor(quality)",
                             "total.sulfur.dioxide")+
  scale_y_continuous(limits=c(0,300))
grid.arrange(qual_so2_scatterplot,qual_so2_boxplot)
```

total.sulfur.dixode also seems to be less in the high score wines than the ones 
with a low score. 

### 10. residual sugar

```{r}
sugar_plot=ggplot(data=whiteWineData,aes(x=residual.sugar))+
  scale_x_continuous(limits=c(0,25))
sugar_hist<-sugar_plot+geom_histogram()
sugar_dens<-sugar_plot+geom_density()
grid.arrange(sugar_hist,sugar_dens,nrow=2)

```

The tail on the right side is long and seems to be right skewed distribution. 
Lot of them have values between 0 to 5

Let us plot the residual.sugar vs quality plots

```{r}
qual_sugar_scatterplot <-f_drawscatter(whiteWineData,"quality","residual.sugar")+
  scale_y_continuous(limits=c(0,20))
qual_sugar_boxplot <-f_drawbox(whiteWineData,"factor(quality)","residual.sugar")+
  scale_y_continuous(limits=c(0,20))
grid.arrange(qual_sugar_scatterplot,qual_sugar_boxplot)

```

residual sugars also seem to have a negative correlation with quality.

Based on the observations till now , alcohol and density seems to have strong 
correlation with quality . Let us also see which other variables have strong 
relationship with density and alcohol and choose for our multivariate analysis.

```{r}
cor(x=whiteWineData[,2:12], y=whiteWineData$density,method="pearson")
```

values show that other than alcohol's strong negative correlation, residual sugar
and total.sulfur.dioxide seem to have a strong positive correlation with density.

```{r}
cor(x=whiteWineData[,2:12], y=whiteWineData$alcohol,method="pearson")
```


For alcohol, there are no strong positive correlation but a lot of negative 
correlation factors( density and other variables that have strong correlation 
with density). 


## Multivariate Analysis

Let us consider density , alcohol , residual.sugar,total.sulfur.dixoide  and 
chlorides for our multivariate analysis.

Starting with density and alcohol vs quality

```{r alcohol_density_quality}

qual_dens_alc_scatplotplus5 <-ggplot(aes(x=density,y=alcohol,color=factor(quality>5)),data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(6,15))

qual_dens_alc_scatplotplus6 <-ggplot(aes(x=density,y=alcohol,color=factor(quality>6)),data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(6,15))

grid.arrange(qual_dens_alc_scatplotplus5,qual_dens_alc_scatplotplus6)

```

Alcohol and Density have a strong negative correlation. We can see that a lot of 
wines that have a quality score greater than 6 are at the top left side and the 
ones less than 6 are concentrated at lower right.


```{r}

whiteWineData.quality_groups<-whiteWineData%>%
  group_by(quality)%>%
  summarise(mean_alc=mean(alcohol),
            mean_den=mean(density),
            mean_sugar=mean(residual.sugar),
            mean_totalso2=mean(total.sulfur.dioxide),
            mean_chlorides=mean(chlorides),
            mean_pH=mean(pH),
            mean_sulph=mean(sulphates),
            n=n()
            )%>%
  arrange(quality)

head(whiteWineData.quality_groups)

```

From the table we can see that 
1. quality has positive correlation with alcohol
2. quality has negative correlation with density,total so2, residual sugars and 
sulphates.

### Quality vs residual sugar and density

```{r sugar_density_quality}

qual_dens_sug_scatplotplus5 <-ggplot(aes(x=density,y=residual.sugar,
                                         color=factor(quality>5)),
                                     data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(0,30))
  
qual_dens_sug_scatplotplus6 <-ggplot(aes(x=density,y=residual.sugar,
                                         color=factor(quality>6)),
                                     data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(0,30))

grid.arrange(qual_dens_sug_scatplotplus5,qual_dens_sug_scatplotplus6)

```

From the plot , we can see that the ones that have quality score greater than 6 
have low residual sugar and density.The ones that have quality score 5 and 6 are 
dispersed across the spectrum.


### Quality vs totalso2 and density
```{r}
qual_dens_totso2_scatplotplus5 <-ggplot(aes(x=density,y=total.sulfur.dioxide,color=factor(quality>5)),
                                        data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(0,300))
  
qual_dens_totso2_scatplotplus6 <-ggplot(aes(x=density,y=total.sulfur.dioxide,color=factor(quality>6)),
                                        data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(0,300))

grid.arrange(qual_dens_totso2_scatplotplus5,qual_dens_totso2_scatplotplus6)


```


This also is inline with the correlation numbers. The ones that have low 
total.sulfur.dioxide and density have quality rating >6


##Final 3 Plots

1. Quality Univariate 
2. Alcohol vs quality and Density vs quality box plots grid
3. Quality vs Alcohol vs Density multivariate

### Final plot1

```{r qual_histogram_FP1}

quality_plot_fp1=ggplot(data=whiteWineData,aes(x=quality))
quality_plot_fp1+geom_histogram(binwidth=1,color=I('black'),fill=I('#099009'),origin=0.5)+
  scale_x_continuous(breaks = seq(2,10,1), lim = c(2,10))+
  xlab("Quality")+
  ylab("No. of Wines")+
  ggtitle("Distribution of Quantity ")

```

Final plot1 Description :

1. Quality has values ranging from 3 to 9
2. Most of the wines have a quality score of 5 and 6 ,6 being highest, very 
  few have the highest of 9 and lowest score of 3.
  

### Final plot 2 

```{r final_plot2}

qual_den_fp2 <-f_drawbox(whiteWineData,"factor(quality)","density")+
  scale_y_continuous(limits=c(0.985,1.005))+
  xlab("Quality")+
  ylab("Density in gm/cm3")+
  ggtitle(" Distribution of Density across quality scores")

qual_alc_fp2 <-f_drawbox(whiteWineData,"factor(quality)","alcohol")+
  xlab("Quality")+
  ylab("Alcohol pH level")+
  ggtitle(" Distribution of Alcohol values across quality scores ")

grid.arrange(qual_den_fp2,qual_alc_fp2)


```

Final plot 2 Description:

1. We can see from the density vs quality plot that the wines of high quality 
  have low density compared to the ones of low score.
2. From the alcohol plot , except for wines of score 3 and 4 , the trend 
  quality score seemed to increase with increase in alcohol level. For 3 and 4 , 
  the total.sulfur.dioxide and residual.sugar were very high which could have 
  contributed to the low score.

###Final plot 3

```{r final_plot_3}


qual_dens_alc_fp3_plus5 <-ggplot(aes(x=density,y=alcohol,
                                     color=factor(quality>5)),
                                 data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(6,15))+
  xlab("Density in gm/cm3")+
  ylab("Alcohol pH level")+
  ggtitle("Alcohol vs Density across different Quality scores >5")+
  scale_colour_discrete(name='Quality',
                        labels=c("Score <=5","Score >5"))
  

qual_dens_alc_fp3_plus6 <-ggplot(aes(x=density,y=alcohol,
                                     color=factor(quality>6)),
                                 data=whiteWineData)+
  geom_jitter(alpha=1/3) +
  geom_smooth(method='lm', aes(group = 1))+
  scale_x_continuous(limits=c(0.985,1.005))+
  scale_y_continuous(limits=c(6,15))+
  xlab("Density in gm/cm3")+
  ylab("Alcohol pH level")+
  ggtitle("Alcohol vs Density across different Quality scores >6 ")+
  scale_colour_discrete(name='Quality',
                        labels=c("Score <=6","Score >6"))

grid.arrange(qual_dens_alc_fp3_plus5,qual_dens_alc_fp3_plus6)


```

Final plot 3 Description :

1. Quality score 6 (which is most common) has dispersed along the density and 
 alcohol axes.
2. But we can clearly see that wines that have quality score greater than 6 
 have high alcohol content and low density.(except for a few exceptions)
  
##Reflection: 

From the entire analysis , we can see that Alcohol content and Density have good 
correlation with the quality score, alcohol being a positive and Density being 
negative. There are other factors like residual.sugar, total.sulfur.dixoide which 
also have a negative correlation with quality score. There was no clear proof 
that these variables are dependent on one another but they definitely contributed 
together towards the quality score.
High values of alcohol , pH and low values of density ,total.sulfur.dixoide and 
residual.sugar seemed to have lead to better quality score.

Although there are trends , there are deviations which I could not explain with the given dataset. For e.g., Density and Alcohol have negative and positive correlation with Quality score in general but wines of Score 3 have greater alcohol content than wines of score 4 and 5. Similarly Wines of score 5 have greater Density than wines of score 3 and 4. I could not find a way to dig deeper to justify these anamolies.

The dataset had wines with quality scores 5 and 6 dominant.Hence we may not be 
able to use this as a training data to predict wines of very high quality. We 
may need more data that has details and composition of high quality wines.